Pool Model - A unified Mathematical\\ Modelling Framework for Food Science

### 2025-09-04 paper/pool_model.tex
reformulate section 3

One on the ways to describe competition between bacteria populations is as a function of limiting resource ratio~\cite{tilman_resource_1977, smith_effects_2002}.
To describe such a situation we consider for simplicity only one common limiting resource pool.

@@

These equations have the structure of a \ac{glv} expanded with different internal states for one species.

@@

Here species $A$ has a longer lag-phase than $B$ but the exponential growth of $A$ is stronger than the one that bacteria $B$ experiences.
As a result, species $A$ was able to suppress species $B$ consuming the larger amount of nutrient and reaching larger bacterial count at the end.

@@

Equations (\ref{eq:LV_simple}) are basically competitive Lotka-Volterra equations that are widely used in ecological studies and show the principle of competitive excursion.
That principle claims that two species in competition for the same resource cannot coexist~\cite{wangersky_lotka-volterra_1978}.

@@

However, except the inhibition due to resource competition, bacteria developed different others active competition mechanisms.
One of the examples of such mechanisms is a production of the antimicrobial compounds, inhibitors or toxins~\cite{wloch-salamon_effect_2008, chao_structured_1981}.

@@

In this way, can suppress the non-preferred species by helping their competitors~\cite{hibbing_bacterial_2010}.

@@

To introduce active competitive mechanisms, we extend our model with one of the species producing the toxic compounds $T$ during its growth:

@@

Where we consider two possible models of the toxin production:

@@

The difference between the two toxin production models for resulting solution of the \acp{ode} is presented in Figure~\ref{fig:2pool_2sp_toxin}.

@@

That means that as long as there are bacteria in the growth phase $G_B>0$ toxin is produced, no matter whether the total bacterial count $N_B$ is still growing.

@@

Such situation with 2 bacteria species, limiting resource and produced inhibitor was modeled, e.g., by Lenski and Hattingh~\cite{lenski_coexistence_1986}.
In our case, we would like to look at rather simplified case reducing the number of variables of the system.

@@

In the first case the inhibition is of minor importance and could be ignored.
In the latter case one can simplify the equations, given $N_A$ and $N_B$ are non-zero:

@@

We introduced the rescaled growth rates $\tilde{\alpha}_A=\alpha_A/(N_tK_B)$ and $\tilde{\alpha}_B=\alpha_B/(N_tK_A)$. Note that $[\tilde{\alpha}]=\mathrm{s}^{-1}$.
Rescaling time with $\tilde{\alpha}_A$ and defining the dimensionless parameter

@@

results in: 

@@

We can separate three cases (which are displayed in Figure~\ref{fig:1pool_2sp_inhibit}):

@@

@@

On the other hand, one can also observe some cooperative behaviors between species in case when it facilitates the competition with others.
It happens if bacteria species as a result of their metabolism produce the compounds that are beneficial for the growth of each other, so-called public goods.
Quite often it happens between the species that are close relatives or clonal populations and such process is called kin selection \cite{west_social_2007}.
Furthermore, neighbors can also benefit from a certain behavior without expressing it by itself, so-called 'social cheaters'~\cite{rainey_evolution_2003}.
The benefit, or mutual activation, of the two species from the presence of the other can be simply introduced by mathematical functions.
For sake of clarity we omit again the lag pool and write:

@@

    \dot{P}_B &= \mathcal{F}_B(R,G_B).

@@

$P_A$ and $P_B$ denote the substance produced by species A and B, respectively.

@@

Setting $\mathcal{B}_A(P_B,R)=\alpha_A P_BR$, $\mathcal{B}_B(P_A,R)=\alpha_B P_A R$, $\mathcal{F}_A(R,G_A)=\kappa_A RG_A$, $\mathcal{F}_B(R,G_B)=\kappa_B RG_B$, rescaling $G_A$, $G_B$, and $P_B$ with $N_t$, $P_A$ with $\kappa_A/\alpha_A$, time with $\alpha_AN_t^2$, defining $\psi=\kappa_A\alpha_B/(\alpha_A^2N_t)$, and $\phi=\kappa_B/(\alpha_A N_t)$ results in:

@@

Here we can see that if $\psi=1$ and $\phi=1$ (see Figure~\ref{fig:1pool_2sp_cooper}a) the production of the public goods of the two species is happening with the same rate as well as the growth of bacteria.
Increasing the production rate of the public good produced by species $B$ (see Figure~\ref{fig:1pool_2sp_cooper}b) causes increasing the growth of species $A$ that profit from the presence of $B$.

>>>>

One of the ways to describe competition between bacterial populations is as a function of the limiting resource ratio~\cite{tilman_resource_1977, smith_effects_2002}.
To this end we consider for simplicity only one common limiting resource pool.

@@

These equations have the structure of a \ac{glv} model expanded with different internal states for one species.

@@

Here species $A$ has a longer lag-phase than $B$ but the exponential growth of $A$ is stronger compared to which is experienced by bacteria $B$.
As a result, species $A$ was able to suppress species $B$, consuminl the larger amount of nutrients and reaching a larger bacterial count at the end.

@@

Equations (\ref{eq:LV_simple}) can be understood as competitive Lotka-Volterra equations that are widely used in ecological studies and show the principle of competitive excursion.
That principle claims that two species which compete for the same resource cannot coexist~\cite{wangersky_lotka-volterra_1978}.

@@

Mutual inhibition between different species can not only result from competition for a shared resource for which bacteria have developed various other active competition mechanisms.
One of the examples of such a mechanisms is the production of antimicrobial compounds, inhibitors or toxins~\cite{wloch-salamon_effect_2008, chao_structured_1981}.

@@

In this way, they can suppress the non-preferred species by helping their competitors~\cite{hibbing_bacterial_2010}.

@@

One direct form of inhibition is the production of toxins.
To introduce active competitive mechanisms, we extend our model with one of the species, producing the toxic compounds $T$ during its growth:

@@

where we consider two possible models of the toxin production

@@

The difference in the solutions of the given \acp{ode} between the two toxin production models is presented in Figure~\ref{fig:2pool_2sp_toxin}.

@@

That means that as long as there are bacteria in the growth phase $G_B>0$, toxin is produced, no matter whether the total bacterial count $N_B$ is still growing.

@@

Such situation with 2 bacteria species, a shared resource and produced inhibitor was modeled, e.g., by Lenski and Hattingh~\cite{lenski_coexistence_1986}.
In our case, we would like to look at a rather simplified case reducing the number of variables of the system.

@@

In the first case the inhibition is of minor importance and could be ignored while in the latter case one can simplify the equations, given $N_A$ and $N_B$ are non-zero:

@@

We introduced the rescaled growth rates $\tilde{\alpha}_A=\alpha_A/(N_tK_B)$ and $\tilde{\alpha}_B=\alpha_B/(N_tK_A)$.
Note that $[\tilde{\alpha}]=\mathrm{s}^{-1}$.
We rescale the time with $\tilde{\alpha}_A$ and define the dimensionless parameter

@@

which results in

@@

We can further separate three cases

@@

which are displayed in Figure~\ref{fig:1pool_2sp_inhibit}.

@@

On the other hand, one can also observe some cooperative behaviors between species in cases when it facilitates the competition with others.
This phenomenon occurs when bacterial metabolism produces compounds that are beneficial for the growth of each other, so-called public goods.
It is most often observed between species which are close relatives or clonal populations.
This process is called kin selection~\cite{west_social_2007}.
Furthermore, neighbors can also benefit from a behavior which does not require any action on an expression level, so-called 'social cheaters'~\cite{rainey_evolution_2003}.
This mutual activation benefits all involved species and can by easily included in our mathematical framework.
For the sake of clarity we omit again the lag pool and write

@@

    \dot{P}_B &= \mathcal{F}_B(R,G_B)

@@

where $P_A$ and $P_B$ denote the substance produced by species A and B, respectively.

@@

Setting $\mathcal{B}_A(P_B,R)=\alpha_A P_BR$, $\mathcal{B}_B(P_A,R)=\alpha_B P_A R$, $\mathcal{F}_A(R,G_A)=\kappa_A RG_A$, $\mathcal{F}_B(R,G_B)=\kappa_B RG_B$, rescaling $G_A$, $G_B$, and $P_B$ with $N_t$, $P_A$ with $\kappa_A/\alpha_A$, time with $\alpha_AN_t^2$, defining $\psi=\kappa_A\alpha_B/(\alpha_A^2N_t)$, and $\phi=\kappa_B/(\alpha_A N_t)$ results in

@@

Here we can see that if $\psi=1$ and $\phi=1$ (see Figure~\ref{fig:1pool_2sp_cooper}a) the production of the public goods of the two species is happening with the same rate as the growth of bacteria.
Increasing the production rate of the public good (see Figure~\ref{fig:1pool_2sp_cooper}b), leads to an increase in the growth of species $A$ that profit from the presence of $B$.
